---
title: "HAEDAT in R"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Simple access to downloading and reading [HAEDAT](https://ipt.iobis.org/hab/resource?r=haedat) data in R

### Requirements

+ [R v4.1+](https://www.r-project.org/)


+ [rlang](https://CRAN.R-project.org/package=rlang)
+ [dplyr](https://CRAN.R-project.org/package=dplyr)
+ [readr](https://CRAN.R-project.org/package=readr)
+ [rappdirs](https://CRAN.R-project.org/package=rappdirs)
+ [finch](https://CRAN.R-project.org/package=finch)
+ [sf](https://CRAN.R-project.org/package=sf)

### Installation

```{r install, eval = FALSE}
remotes::install_github("BigelowLab/haedat")
```

### Initial Usage


#### Establish a root data directory

Please establish the storage directory with `haedat_root()`. You can store data where ever you would like, but we suggest that you accept the default and ignore of all of details and mayhem that ensue with custom installations.

```{r root}
library(haedat)
haedat::haedat_root()
```

#### Fetch the data

The following will fetch and store the haedat data in DWCA format (Darwin Core Archive) from [OBIS](https://ipt.iobis.org/hab/resource?r=haedat).  You only need to do this occasionally as the data is updated online.
```{r fetch_obis,  eval = FALSE}
fetch_haedat(where = "obis")
```

You can also download the data as a single CSV from [IODE](http://haedat.iode.org/advancedSearch.php).  Like the DWCA, you only need to 
refresh this occasionally.

```{r fetch_iode,  eval = FALSE}
fetch_haedat(where = "iode")
```

### List your data

Listings of locally stored data are returned in reverse sort order (since the files
are version named).

```{r listing}

list_haedat(source = "obis")

list_haedat(source = "iode")

list_haedat(source = "all")
```

### Usage


### OBIS data as DWCA

Read the OBIS data into a list or three elements: `events`, `occurrences` and `extendedmeasurementorfact`.  Each of the tables has an `id` which can be used for merging.

```{r read_obis}
suppressPackageStartupMessages({
  library(haedat)
  library(dplyr)
  library(sf)
  library(maps)
})

obis <- read_haedat(source = "obis")
lapply(obis, head)
```

### event

Let's take a look at the spatial distribution.

```{r plot_obis}
event <- obis$event
maps::map()
plot(sf::st_geometry(event),
     col = "orange",
     add = TRUE)
```

What about the distribution through time?  Hmmm.  The `eventDate` variable is character which gives one that 'oh, dear' feeling.  Let's see about how long the strings are; if they are all the same it should be easy to convert to a usable Date or numeric class.

```{r eventDate-length}
lens <- nchar(event$eventDate)
tlens <- table(lens)
tlens
```
Oh, nuts!  What do some of the different lengths look like?

```{r eventDate-explore}
sapply(names(tlens),
       function(len){
         ix <- lens == as.numeric(len)
         sprintf("len=%s: %s", as.character(len),
                     paste(head(event$eventDate[ix]), collapse = ", "))
       })
```

Groan.

Maybe we just take the one with 4-digits in the first four positions, call that `eventYear` and drop what doesn't convert.

```{r eventYear_obis}
event <- dplyr::mutate(event,
                       eventYear = as.numeric(substring(eventDate, 1,4))) |>
  dplyr::filter(!is.na(eventYear) & eventYear > 1990)
hist(event$eventYear, main = "OBIS HAEDAT events since 1990",
     xlab = 'Year',
     breaks = seq(from = min(event$eventYear), to = max(event$eventYear)))
```


### IODE data as CSV

```{r read_iode}
iode <- read_haedat(source = "iode")
glimpse(iode)
```


```{r plot_iode}
maps::map()
plot(sf::st_geometry(iode),
     col = "orange",
     add = TRUE)
```

```{r eventYear_iode}

hist(dplyr::filter(iode, !is.na(eventYear) & eventYear > 1990) |>
       dplyr::pull(eventYear), 
     main = "IODE HAEDAT events since 1990",
     xlab = 'Year',
     breaks = seq(from = 1990, to = max(iode$eventYear)))
```


### Comparing

```{r}
obis <- read_haedat(source = "obis", form = "tibble")
iode <- read_haedat(source = "iode", form = "tibble")
dim(obis$event)
dim(iode)
```
