---
title: "HAEDAT in R"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Simple access to downloading and reading [HAEDAT](https://ipt.iobis.org/hab/resource?r=haedat) data in R

### Requirements

+ [R v4.1+](https://www.r-project.org/)


+ [rlang](https://CRAN.R-project.org/package=rlang)
+ [dplyr](https://CRAN.R-project.org/package=dplyr)
+ [readr](https://CRAN.R-project.org/package=readr)
+ [rappdirs](https://CRAN.R-project.org/package=rappdirs)
+ [finch](https://CRAN.R-project.org/package=finch)
+ [sf](https://CRAN.R-project.org/package=sf)

### Installation

```{r install, eval = FALSE}
remotes::install_github("BigelowLab/haedat")
```

### Initial Usage


#### Establish a root data directory

Please establish the storage directory with `haedat_root()`. You can store data where ever you would like, but we suggest that you accept the default and ignore of all of details and mayhem that ensue with custom installations.

```{r root, eval = FALSE}
haedat::haedat_root()
```

#### Fetch the data

The following will fetch and store the haedat data in DWCA format (Darwin Core Archive).  You only need to do this occasionally as the data is updated online.
```{r fetch,  eval = FALSE}
headat::fetch_haedat()
```


#### Usage

Read the data into a list or three elements: `events`, `occurrences` and `extendedmeasurementorfact`.  Each of the tables has an `id` which can be used for merging.

```{r read}
suppressPackageStartupMessages({
  library(haedat)
  library(dplyr)
  library(sf)
  library(maps)
})

x <- read_haedat()
lapply(x, head)
```


### event

Let's take a look at the spatial distribution.

```{r plot}
event <- x$event
maps::map()
plot(sf::st_geometry(event),
     col = "orange",
     add = TRUE)
```

What about the distribution through time?  Hmmm.  The `eventDate` variable is character which gives one that 'oh, dear' feeling.  Let's see about how long the strings are; if they are all the same it should be easy to convert to a usable Date or numeric class.

```{r eventDate-length}
lens <- nchar(event$eventDate)
tlens <- table(lens)
tlens
```
Oh, nuts!  What do some of the different lengths look like?

```{r eventDate-explore}
sapply(names(tlens),
       function(len){
         ix <- lens == as.numeric(len)
         sprintf("len=%s: %s", as.character(len),
                     paste(head(event$eventDate[ix]), collapse = ", "))
       })
```

Groan.

Maybe we just take the one with 4-digits in the first four positions, call that `eventYear` and drop what doesn't convert.

```{r eventYear}
event <- dplyr::mutate(event,
                       eventYear = as.numeric(substring(eventDate, 1,4))) |>
  dplyr::filter(!is.na(eventYear) & eventYear > 1990)
hist(event$eventYear, main = "HAEDAT events since 1990",
     xlab = 'Year',
     breaks = seq(from = min(event$eventYear), to = max(event$eventYear)))
```
